---
title: "Fitting models H2"
output: html_document
date: "2025-09-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load image

```{r}
data.path = dirname(getwd())
```

Load libraries:
```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(purrr)
library(minpack.lm)
library(ggplot2)
library(patchwork)
library(scales)
```

## MODELS
### Define functions
```{r}
# Yashiro/Smith-Downey
# g_T_smith_downey = function(T) {
  #return(1 / (1 + exp(-0.1718 * (T + 273.15) + 46.9388)))
#}

#f_M_smith_downey <- function(frac_water_pores_avg) {
  #f_M <- 14.3 * frac_water_pores_avg - 1.14
  #f_M <- pmin(pmax(f_M, 0), 1)
  #return(f_M)
#}

# Ehhalt
# temp function
g_T <- function(T) {
  return(1 / (1 + exp(-(T - 3.8) / 6.7)) + 1 / (1 + exp((T - 62.7) / 7.1)) - 1)
}

# two different functions for moisture dependence of sand and loam
f_M_sand <- function(frac_water_pores_avg) {
  f_M <- 0.00936 * (frac_water_pores_avg - 0.0264) * (1 - frac_water_pores_avg) / 
         (frac_water_pores_avg^2 - 0.1715 * frac_water_pores_avg + 0.03144)
  f_M <- pmax(f_M, 0)
  return(f_M)
}

f_M_loam <- function(frac_water_pores_avg) {
  f_M <- 0.01997 * (frac_water_pores_avg - 0.05369) * (0.8508 - frac_water_pores_avg) / 
         (frac_water_pores_avg^2 - 0.7541 * frac_water_pores_avg + 0.2806)
  f_M <- pmax(f_M, 0)
  return(f_M)
}

# A few moisture parameterization and generic wrapper

# Moisture functions
f_M_bertagni <- function(s, s_ws, beta1, beta2, N) {
  if (s > s_ws && s < 1) {
    f_M <- 1 / N * (s - s_ws)^beta1 * (1 - s)^beta2
  } else {
    f_M <- 0
  }
  return(f_M)
}


# For fitting Matteo's model
calc_k_psi <- function(beta1, psi_ws, psi_opt, prek, obs) {
  ot <- obs
  
  ot$s_ws <- mapply(function(psi_MPa, b) (psi_MPa / psi_ws)^(1 / b), ot$Psi_MPa, ot$b)
  ot$s_opt <- mapply(function(psi_MPa, b) (psi_MPa / psi_opt)^(1 / b), ot$Psi_MPa, ot$b)
  
  ot$beta1 <- beta1
  ot$beta2 <- mapply(function(beta1, s_opt, s_ws) beta1 * (1 - s_opt) / (s_opt - s_ws), 
                     ot$beta1, ot$s_opt, ot$s_ws)
  ot$N <- mapply(function(s_opt, s_ws, beta1, beta2) (s_opt - s_ws)^beta1 * (1 - s_opt)^beta2, 
                 ot$s_opt, ot$s_ws, ot$beta1, ot$beta2)
  
  ot$moisture <- mapply(psi_to_moisture, ot$WP, ot$Psi_MPa, ot$b)
  ot$predicted_k <- mapply(function(moisture, s_ws, beta1, beta2, N, t_ref) 
                            prek * f_M_bertagni(moisture, s_ws, beta1, beta2, N) * g_T(t_ref), 
                            ot$moisture, ot$s_ws, ot$beta1, ot$beta2, ot$N, MoreArgs = list(t_ref = t_ref))
  
  return(ot)
}

calc_k_moisture <- function(beta1, s_ws, s_opt, prek, obs) {
  ot <- obs
  
  ot$beta1 <- beta1
  ot$beta2 <- mapply(function(beta1, s_opt, s_ws) beta1 * (1 - s_opt) / (s_opt - s_ws), 
                     ot$beta1, s_opt, s_ws)
  ot$N <- mapply(function(s_opt, s_ws, beta1, beta2) (s_opt - s_ws)^beta1 * (1 - s_opt)^beta2, 
                 s_opt, s_ws, ot$beta1, ot$beta2)
  
  ot$predicted_k <- mapply(function(moisture_for_fit, s_ws, beta1, beta2, N, t_ref) 
                            prek * f_M_bertagni(moisture_for_fit, s_ws, beta1, beta2, N) * g_T(t_ref), 
                            ot$moisture_for_fit, s_ws, ot$beta1, ot$beta2, ot$N, MoreArgs = list(t_ref = t_ref))
  
  return(ot)
}

# Function-dependent fit
calc_k_psi_g <- function(fM, param_df, obs, t_ref) {
  ot <- obs
  
  if (fM == 'Bertagni') {
    ot$s_ws <- mapply(function(psi_MPa, b) (psi_MPa / param_df$psi_ws)^(1 / b), ot$Psi_MPa, ot$b)
    ot$s_opt <- mapply(function(psi_MPa, b) (psi_MPa / param_df$psi_opt)^(1 / b), ot$Psi_MPa, ot$b)
    
    ot$beta2 <- mapply(function(s_opt, s_ws) param_df$beta1 * (1 - s_opt) / (s_opt - s_ws), 
                       ot$s_opt, ot$s_ws)
    ot$N <- mapply(function(s_opt, s_ws, beta1, beta2) (s_opt - s_ws)^beta1 * (1 - s_opt)^beta2, 
                   ot$s_opt, ot$s_ws, param_df$beta1, ot$beta2)
    
    ot$predicted_k <- mapply(function(moisture_for_fit, s_ws, beta1, beta2, N, soil_type, t_ref) 
                              param_df[[paste0("km_", soil_type)]] * f_M_bertagni(moisture_for_fit, s_ws, beta1, beta2, N) * g_T(t_ref), 
                              ot$moisture_for_fit, ot$s_ws, param_df$beta1, ot$beta2, ot$N, ot$soil_type, MoreArgs = list(t_ref = t_ref))
    
  } else if (fM == 'Sun') {
    ot$s_opt <- mapply(function(psi_MPa, b) (psi_MPa / param_df$psi_opt * 1e-3)^(1 / b), ot$Psi_MPa, ot$b)
    ot$s_ws <- mapply(function(psi_MPa, b) (psi_MPa / param_df$psi_ws * 1e-3)^(1 / b), ot$Psi_MPa, ot$b)
    
    ot$predicted_k <- mapply(function(moisture_for_fit, s_ws, s_opt, soil_type, t_ref) 
                              param_df[[paste0("km_", soil_type)]] * f_M_Sun(moisture_for_fit, param_df$beta1, s_ws, s_opt) * g_T(t_ref), 
                              ot$moisture_for_fit, ot$s_ws, ot$s_opt, ot$soil_type, MoreArgs = list(t_ref = t_ref))
    
  } else if (fM == 'Kesselmeier') {
    ot$s_opt <- mapply(function(psi_MPa, b) (psi_MPa / param_df$psi_opt * 1e-3)^(1 / b), ot$Psi_MPa, ot$b)
    ot$s_ws <- mapply(function(psi_MPa, b) (psi_MPa / param_df$psi_ws * 1e-3)^(1 / b), ot$Psi_MPa, ot$b)
    
    ot$predicted_k <- mapply(function(moisture_for_fit, s_ws, s_opt, k_w1, k_w2, soil_type, t_ref) 
                              param_df[[paste0("km_", soil_type)]] * f_M_Kesselmeier(moisture_for_fit, k_w1, k_w2, s_ws, s_opt) * g_T(t_ref), 
                              ot$moisture_for_fit, ot$s_ws, ot$s_opt, param_df$k_w1, param_df$k_w2, ot$soil_type, MoreArgs = list(t_ref = t_ref))
  }
  
  return(ot)
}

```



## Read in data
```{r, echo=FALSE}
data.path = dirname(getwd())

# observations
dat = read.table(paste0(data.path, "/data/H2_experiments_allData_finalCorrected.txt"), header=TRUE, sep="\t", check.names = FALSE)

# soil properties
soil_properties = read.csv(paste0(data.path, "/data/Bertagni.csv"), row.names = 1)
colnames(soil_properties) = c("Psi..kPa", "b", "n", "sws", "sopt")
```

## Calculate Theta_w

### water (g) per gram of dry soil (i.e., water lost upon drying/(total soil - water lost upon drying))
#### soil_g: Mean bulk soil weight in experiment (g)

#### thetaW ~ volumetric moisture per dry soil
#### thetaW/thetaP ~ % saturation (thetaW/porosity)

```{r}
dat = dat %>%
  mutate(#thetaW = water_perDrySoil* (soil_g/bulk_vol_soil_ml),
        thetaW = water_perDrySoil*bulk_density_gPercm3,
         s = thetaW/porosity)
```

## soil moisture prep
### moisture-psi interconversion functions
```{r}
# Psi <-> s

psi_to_moisture <- function(psi,psi_ref,b) {
  s = (psi / psi_ref)^(-1 / b)
  return(s)
}

moisture_to_psi <- function(s,psi_ref,b) {
    psi = psi_ref * s^(-b)
    return(psi)
}
```

## adjust s based on literature values
```{r}
# Define a mapping from soil types to USDA soil classes
soil_map <- c(Forest = "Sandy Loam", Meadow = "Silt Loam", PB_sand = "Loamy Sand")
# Reverse mapping
rsoil_map <- setNames(names(soil_map), soil_map)

soil_properties = soil_properties %>% 
              mutate(Psi_MPa = `Psi..kPa` / 1e3) %>%  # Convert Psi to MPa 
              rownames_to_column(var = "soil_type") %>%  # Reset index
              mutate(soil_type = rsoil_map[soil_type]) %>%  # Map soil types
              drop_na() %>%  # Remove NA values
              column_to_rownames(var = "soil_type")
soil_properties

# estimate psi_ref, b

# Define function to estimate soil moisture given psi
psi_moisture_opt = function(psi, psi_ref, b) {
  s = psi_to_moisture(psi, psi_ref, b) - psi_to_moisture(min(psi), psi_ref, b)
  return(s)
}
 
## deep copy of dat
obs = dat

# Filter observations: only use WP between -70 and -0.5
obsf = obs %>%
  filter(WP < -0.5, WP > -70)

# Group by soil_type
groups = split(obsf, obsf$soil_type)

# Initialize a list to store parameters for each soil type
exp_soil_properties = list()

for (name in names(groups)) {
  group = groups[[name]]
  fit = nls(s - min(s) ~ psi_moisture_opt(WP, psi_ref, b), 
             data = group,
             start = list(psi_ref = -0.1, b = 0.6),
             lower = c(psi_ref = -1, b = 0),
             upper = c(psi_ref = -1e-9, b = 10),
             algorithm = "port")
  params <- coef(fit)
  #exp_soil_properties[[name]] <- c(params["psi_ref"], params["b"])
  exp_soil_properties[[name]] <- tibble(soil_type = name, Psi_MPa = params["psi_ref"], b = params["b"])
}

# Convert exp_soil_properties from a list of tibbles to a single tibble
exp_soil_properties_df <- bind_rows(exp_soil_properties)
exp_soil_properties_df

# Merge estimated parameters with observed data
obs = obs %>%
  left_join(exp_soil_properties_df, by = "soil_type") %>%
  mutate(s_from_WP = psi_to_moisture(WP, Psi_MPa, b))

# Compute minimum water values per soil type
min_water = obs %>%
  group_by(soil_type) %>%
  summarize(across(c(s_from_WP, s), min, .names = "min_{.col}"), .groups = "drop")

# Apply correction to s values
obs = obs %>%
  left_join(min_water, by = "soil_type") %>%
  mutate(s_corrected = s - min_s + min_s_from_WP) %>%
  select(-starts_with("min_"))

min_water
obs %>% select(s, s_from_WP, s_corrected) %>% unique()

```

## plot soil water characteristic curves
```{r}
# Define the soil types to plot
soil_types = c("PB_sand", "Meadow", "Forest")

# Create plots for each soil type
plots = map(soil_types, function(soil) {
  
  # Filter data for the given soil type
  O = obs %>% filter(soil_type == soil)
  
  # Scatter plot of log-transformed WP vs. s
  p1 = ggplot(O, aes(x = log10(-WP), y = log10(s))) +
    geom_point() +
    ggtitle(soil) +
    theme_minimal()
  
  # Filter further for reliable WP range
  Of = O %>% filter(WP < -0.5, WP > -150)
  
  # Scatter plot of log-transformed WP vs. s and s_corrected
  p2 = ggplot(Of, aes(x = log10(-WP))) +
    geom_point(aes(y = log10(s), color = "s")) +
    geom_point(aes(y = log10(s_corrected), color = "s (corrected)")) +
    labs(x = "-log10(WP)", y = "log10(s)", color = "Legend") +
    theme_minimal()
  
  # Return combined plots
  p1 / p2  # Stack the two plots vertically
})

# Combine all plots into a grid
wrap_plots(plots, ncol = 3) 
```

## Fit data
```{r}
# Define the first-order function
first_order = function(t, k, H2_0) {
  return(H2_0 * exp(-k * t))
}

# Fit function
fit_generic = function(x, y, ftype) {
  if (ftype == 'first_order') {
    # Fit the first-order model
    fit <- tryCatch({
      nls(y ~ H2_0 * exp(-k * x), 
          start = list(k = 0.01, H2_0 = 0.6),
          lower = c(0, 0),
          upper = c(Inf, Inf),
          algorithm = "port")
    }, error = function(e) return(NULL))
    
    if (!is.null(fit)) {
      params = coef(fit)
      return(data.frame(k = params['k'], H2_0 = params['H2_0']))
    }
  }
  return(NULL)
}

# Apply fitting function to the data
Out_f = obs %>%
  group_by(soil_type, sample_id, replicate) %>%
  do({
    fit_generic(.$time_min, .$h2_ppmv_dil_corr, 'first_order')
  }) %>%
  ungroup()

# Rename columns
Out_f = Out_f %>%
  rename(k_f = k, H2_0_f = H2_0)

# Merge with observed data at time_min == 0
O = obs %>% filter(time_min == 0) %>%
  select(soil_type, sample_id, replicate, h2_ppmv_dil_corr, WP, vol_HS_ml, bulk_vol_soil_ml) #************

OC = left_join(Out_f, O, by = c('soil_type', 'sample_id', 'replicate'))

# Add new calculated columns
OC = OC %>%
  mutate(k_f_V = k_f * vol_HS_ml / bulk_vol_soil_ml * 1/60)

# Summary statistics (mean, standard deviation, SEM)
#OC_g = OC %>%
  #group_by(soil_type, sample_id) %>%
  #select_if(is.numeric) %>%
  #summarise(across(everything(), list(mean = mean, std = sd, sem = ~sd(.)/sqrt(n())), .names = "{.col}_{.fn}"))

OC_g = OC %>%
  group_by(soil_type, sample_id) %>%
  mutate(WP_mean = mean(WP),
         WP_std = sd(WP),
         WP_sem = WP_std/sqrt(n()),
         k_f_V_mean = mean(k_f_V),
         k_f_V_std = sd(k_f_V),
         k_f_V_sem = k_f_V_std/sqrt(n())
         )

# Print the result
# print(OC_g)
```


# plot data and model outputs
```{r}
# Prepare Data for Plotting
moisture <- exp(seq(log(1e-5), 0, length.out = 1000))
t_ref <- 22 # in C
k_to_compare = 'k_f_V'
WP_to_use = 'WP'
s_to_use = 's_corrected'
max_WP = -.25

# convert moisture to psi for each soil type 
psi = list()
for (st in soil_types) {
  psi[[st]] <- moisture_to_psi(
    moisture,
    exp_soil_properties_df %>% filter(soil_type == st) %>% pull(Psi_MPa),
    exp_soil_properties_df %>% filter(soil_type == st) %>% pull(b)
  )
}

# Ehhalt model
# ck_ehhalt = list()
#ck_ehhalt$PB_sand = data.frame(predicted_k = f_M_sand(moisture) * g_T(t_ref))
#ck_ehhalt$Forest = data.frame(predicted_k = f_M_loam(moisture) * g_T(t_ref))

## convert Ehhalt output to dataframe
# df_ehhalt = map_dfr(ck_ehhalt, ~ as.data.frame(.x) %>% mutate(model = "Ehhalt"), .id = "soil_type")
#ls_ehhalt = list()
#for (st in soil_types[-2]) {
#  ls_ehhalt[[st]] = data.frame("model" = "Ehhalt", "psi" = psi[[st]], "predicted_k" = #ck_ehhalt[[st]]$predicted_k)
# }
# df_ehhalt = map_dfr(ls_ehhalt, ~ as.data.frame(.x), .id="soil_type") %>%
#  group_by(soil_type) %>%
#  mutate(pred_k_scaled = predicted_k/max(predicted_k))

# Smith-downey model
# ck_smith_downey = list()
# for (st in soil_types) {
#  ck_smith_downey[[st]] = data.frame (moisture = moisture,
#  predicted_k = 0.1227 * f_M_smith_downey(moisture) * g_T_smith_downey(t_ref),
#  ref = 'Smith_Downey')
# }

## convert Smith-Downey output to dataframe
# ls_smith_downey = list()
# for (st in soil_types) {
#  ls_smith_downey[[st]] = data.frame("model" = "Smith-Downey", "psi" = psi[[st]], "predicted_k" = # ck_smith_downey[[st]]$predicted_k)
# }
# df_smith_downey = map_dfr(ls_smith_downey, ~ as.data.frame(.x), .id="soil_type") %>%
#  group_by(soil_type) %>%
#  mutate(pred_k_scaled = predicted_k/max(predicted_k))

# bertagni models
param_bertagni <- data.frame(
  beta1 = 0.4, psi_ws = -3, psi_opt = -0.3,
  km_Forest = 0.03, km_Meadow = 0.03, km_PB_sand = 0.03
)
param_bertagni_new <- data.frame(
  beta1 = 1, psi_ws = -100, psi_opt = -0.5,
  km_Forest = 0.06, km_Meadow = 0.06, km_PB_sand = 0.06
)
ck_bertagni <- list()
ck_bertagni_new <- list()

### Prepare Data for k Calculation
D = data.frame(moisture_for_fit = moisture)
for (st in c('Forest', 'Meadow', 'PB_sand')) {
  D$soil_type = st
  D$Psi_MPa = exp_soil_properties_df %>% 
    filter(soil_type == st) %>% 
    pull(Psi_MPa)
  D$b = exp_soil_properties_df %>% 
    filter(soil_type == st) %>% 
    pull(b)
  
  ck_bertagni[[st]] = calc_k_psi_g('Bertagni', param_bertagni, D, t_ref)
  ck_bertagni_new[[st]] = calc_k_psi_g('Bertagni', param_bertagni_new, D, t_ref)
}

## convert bertagni outputs to dataframes
ls_bertagni = list()
for (st in soil_types) {
  ls_bertagni[[st]] = data.frame("model" = "-3 MPa", "psi" = psi[[st]], "predicted_k" = ck_bertagni[[st]]$predicted_k)
}
df_bertagni = map_dfr(ls_bertagni, ~ as.data.frame(.x), .id="soil_type") %>%
  group_by(soil_type) %>%
  mutate(pred_k_scaled = predicted_k/max(predicted_k))

ls_bertagni_new = list()
for (st in soil_types) {
  ls_bertagni_new[[st]] = data.frame("model" = "-100 MPa", "psi" = psi[[st]], "predicted_k" = ck_bertagni_new[[st]]$predicted_k)
}
df_bertagni_new = map_dfr(ls_bertagni_new, ~ as.data.frame(.x), .id="soil_type") %>%
  group_by(soil_type) %>%
  mutate(pred_k_scaled = predicted_k/max(predicted_k))
```


## Plot data and model predictions

```{r}
# function for symlog transformation
symlog_trans <- function(threshold = 1) {
  trans_new("symlog",
    transform = function(x) sign(x) * log10(1 + abs(x / threshold)),
    inverse = function(x) sign(x) * threshold * (10^abs(x) - 1),
    breaks = extended_breaks(),
    domain = c(-Inf, Inf)
  )
}

# plot

OC_g = OC_g %>%
  group_by(soil_type) %>%
  mutate(k_f_V_perc = 100*k_f_V/max(k_f_V)) %>%
  mutate(k_f_V_scaled = k_f_V/max(k_f_V))

pdf(file =  paste0(data.path, "/data/Figure_1b_corr.pdf"), width = 6, height = 4)
ggplot(data = OC_g, aes(x = WP, y = k_f_V_scaled)) +
  geom_line(data = df_bertagni, aes(x=psi, y=pred_k_scaled, color=soil_type, linetype = model), linewidth = 1) +
  geom_line(data = df_bertagni_new, aes(x=psi, y=pred_k_scaled, color=soil_type, linetype = model), linewidth = 1) +
  #geom_line(data= df_ehhalt, aes(x=psi, y=pred_k_scaled, color=soil_type, linetype = model), linewidth = 1) +
  #geom_line(data= df_smith_downey, aes(x=psi, y=pred_k_scaled, color=soil_type, linetype = model), linewidth = 1) +
  geom_point(aes(fill=soil_type), size=2.5, shape=21, color="black", stroke=0.3) +
  scale_x_continuous(trans = symlog_trans(threshold = 1), limits = c(-150, -0.5), breaks = c(-100, -50, -10, -1, -0.01), expand = c(0,0)) +
  scale_fill_manual(values = c("forestgreen", "gold", "violetred")) +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  scale_linetype_manual(values = c("dotdash", "solid",  "dotted", "longdash")) +
  xlab(expression(psi)) +
  theme_minimal() +
  theme(axis.text = element_text(color="black", size=12))
dev.off()
```

# save image
```{r}
data.path
#save.image(paste0(data.path, "/data/fitting_models_updatedCorr.RData"))
```


