---
title: "H2_uptakeRates_final"
output: html_document
date: "2025-11-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Load libraries:
```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(ggpmisc)
library(patchwork)
```

### Read in data:

```{r, echo=FALSE}
data.path = dirname(getwd())
dat = read.table(paste0(data.path, "/data/H2_experiments_allData_finalCorrected.txt"), header=TRUE, sep="\t", check.names = FALSE)
```

### Calculate percent saturation as volumetric moisture% * porosity:

```{r}
dat = dat %>%
  mutate(percSat = volumetric_moisture_perDrySoil_rhoW/porosity)
```

### Filter out air-dried samples before estimating uptake rates
```{r}
dat_ed = dat %>%  
  filter(sample_id != "air-dried")

# select variables of interest
dat_ed = dat_ed %>%
  select(soil_type, sample_label, sample_id, replicate, soil_g, measured_percS, WP, volumetric_moisture_perDrySoil_rhoW, percSat, time_min, h2_ppmv_dil_corr)

```

### Define and apply an exponential decay function:
```{r}
uptake.fun <- function(t, init, k) (init*exp(-k*t))
# k = bacterial uptake rate; t = time in min; init = starting concentration

# apply function to each soil type separately:
dat.models = dat_ed %>%
  group_by(soil_type, sample_id, replicate, soil_g, measured_percS, WP, volumetric_moisture_perDrySoil_rhoW, percSat) %>% nest() %>%
  mutate(model = purrr::map(data, ~nls(h2_ppmv_dil_corr ~ uptake.fun(time_min, h2_ppmv_dil_corr[1], k1), data=., start=list(k1=0.001))),
         model_summaries = map(model, broom::glance),
         model_coef = map(model, broom::tidy))

# get a df with estimated k values
k_estimates = dat.models %>%  
  unnest(model_coef, .drop = TRUE) %>%
  select(soil_type, sample_id, replicate, soil_g, measured_percS, WP, volumetric_moisture_perDrySoil_rhoW, percSat, estimate, std.error)

# combine k estimates and the initial df for plotting

# get initial H2 concentrations
init_ppms = dat_ed %>%
  group_by(soil_type, sample_id, replicate) %>%
  summarise(init = first(h2_ppmv_dil_corr))

# combine initial concentrations with rate constant estimates
dat.k.init = left_join(k_estimates, init_ppms)

# add rate constant estimates and initial concentrations to the original df
dat.combined = left_join(dat_ed, dat.k.init)
```


### estimate curves and plot each soil type separately: 
```{r}
# calculate mean %S for plotting replicates together:
dat.combined = dat.combined %>%
  group_by(soil_type, sample_label) %>%
  mutate(mean_percSat = round(100*mean(percSat), 1))

# forest soil curves
forest_curves = dat.combined %>%
  filter(soil_type == "Forest") %>%
  group_by(sample_id, replicate) %>%
  mutate(pred = uptake.fun(time_min, init, estimate)) %>%
  ggplot(aes(x=time_min, color = factor(replicate), label = paste0("-k = ", round(estimate,3)))) +
  geom_point(aes(y=h2_ppmv_dil_corr), size = 1.5) +
  facet_wrap(~mean_percSat, scales = "free_x", ncol = 2) +
  geom_text(data = (dat.combined %>% filter(soil_type == "Forest" & replicate == 1)), x=45, y=0.5, check_overlap = TRUE) +
  geom_text(data = (dat.combined %>% filter(soil_type == "Forest" & replicate == 2)), x=45, y=0.5, check_overlap = TRUE, vjust=2) +
  geom_text(data = (dat.combined %>% filter(soil_type == "Forest" & replicate == 3)), x=45, y=0.5, check_overlap = TRUE, vjust=4) +
  # or use geom_line but gives weird looking line.
  # geom_line(aes(y= pred, group = replicate), size=0.5) +
  geom_smooth(aes(y= pred, group = replicate), size=0.5, se=FALSE) +
  theme_linedraw() +
  #scale_color_manual(values = c("#118dd5", "#d5118d", "#62940c")) +
    theme(strip.text = element_text(size = 14),
        axis.title = element_text(color="black", size=12),
        axis.text = element_text(color="black", size=11)) +
  xlab("Time (min)") +
  ylab("Headspace H2 ppmv") +
  ggtitle("Forest")

# meadow soil curves
meadow_curves = dat.combined %>%
  filter(soil_type == "Meadow") %>%
  group_by(sample_id, replicate) %>%
  mutate(pred = uptake.fun(time_min, init, estimate)) %>%
  ggplot(aes(x=time_min, color = factor(replicate), label = paste0("-k = ", round(estimate,3)))) +
  geom_point(aes(y=h2_ppmv_dil_corr), size = 1.5) +
  facet_wrap(~mean_percSat, scales = "free_x", ncol = 2) +
  geom_text(data = (dat.combined %>% filter(soil_type == "Meadow" & replicate == 1)), x=45, y=0.5, check_overlap = TRUE) +
  geom_text(data = (dat.combined %>% filter(soil_type == "Meadow" & replicate == 2)), x=45, y=0.5, check_overlap = TRUE, vjust=2) +
  geom_text(data = (dat.combined %>% filter(soil_type == "Meadow" & replicate == 3)), x=45, y=0.5, check_overlap = TRUE, vjust=4) +
  geom_smooth(aes(y= pred, group = replicate), size=0.5, se=FALSE) +
  theme_linedraw() +
  #scale_color_manual(values = c("#118dd5", "#d5118d", "#62940c")) +
    theme(strip.text = element_text(size = 14),
        axis.title = element_text(color="black", size=12),
        axis.text = element_text(color="black", size=11)) +
  xlab("Time (min)") +
  ylab("Headspace H2 ppmv") +
  ggtitle("Meadow")

# PB sand curves
pb_curves = dat.combined %>% filter(sample_id != 1.5) %>%
  filter(soil_type == "PB_sand") %>%
  group_by(sample_id, replicate) %>%
  mutate(pred = uptake.fun(time_min, init, estimate)) %>%
  ggplot(aes(x=time_min, color = factor(replicate), label = paste0("-k = ", round(estimate,3)))) +
  geom_point(aes(y=h2_ppmv_dil_corr), size = 1.5) +
  facet_wrap(~mean_percSat, scales = "free_x", ncol = 3) +
  geom_text(data = (dat.combined %>% filter(sample_id != 1.5 & soil_type == "PB_sand" & replicate == 1)), x=45, y=0.5, check_overlap = TRUE) +
  geom_text(data = (dat.combined %>% filter(sample_id != 1.5 & soil_type == "PB_sand" & replicate == 2)), x=45, y=0.5, check_overlap = TRUE, vjust=2) +
  geom_text(data = (dat.combined %>% filter(sample_id != 1.5 & soil_type == "PB_sand" & replicate == 3)), x=45, y=0.5, check_overlap = TRUE, vjust=4) +
  # or use geom_line but gives weird looking line.
  geom_smooth(aes(y= pred, group = replicate), size=0.5, se=FALSE) +
  theme_linedraw() +
  #scale_color_manual(values = c("#118dd5", "#d5118d", "#62940c")) +
  theme(strip.text = element_text(size = 14),
        axis.title = element_text(color="black", size=12),
        axis.text = element_text(color="black", size=11)) +
  xlab("Time (min)") +
  ylab("Headspace H2 ppmv") +
  ggtitle("PB sand")
  
# fig. S7
(forest_curves + theme(legend.position = "none") )+
  (meadow_curves + theme(legend.position = "none")) + pb_curves
```


### prep new df to include air-dried samples
```{r}
# new df with air-dried samples, set estimate to 0
df_airdried = dat %>% filter(sample_id == "air-dried") %>%
  select(soil_type, sample_id, replicate, soil_g, measured_percS, WP, volumetric_moisture_perDrySoil_rhoW, percSat, time_min, h2_ppmv_dil_corr) %>%
  mutate(estimate = 0,
         std.error = 0)

# calculate mean %S for plotting replicates together:
df_airdried = df_airdried %>%
  group_by(soil_type, sample_id) %>%
  mutate(mean_percSat = round(100*mean(percSat), 1))

# extract initial H2 concentrations into a new column
init_airdried = df_airdried %>%  
  group_by(soil_type, replicate) %>%
  summarise(init = first(h2_ppmv_dil_corr))

df_airdried = left_join(df_airdried, init_airdried)

dat.combined %>% select(soil_type, replicate, percSat) %>% 
  filter(soil_type == "Meadow") %>%unique()
```

### plot air-dried measurements
```{r}
# estimated measurement error (standard error in the mean) = 0.17 ppm
# fig. S2
df_airdried %>%
  filter(sample_id == "air-dried") %>%
  ggplot(aes(x=time_min, y = h2_ppmv_dil_corr)) +
  geom_point(aes(colour = as.factor(replicate))) +
  geom_errorbar(aes(ymin = h2_ppmv_dil_corr-0.17, ymax = h2_ppmv_dil_corr+0.17, width = 0.002), color="lightgray") +
  facet_grid(~mean_percSat, scales = "free") +
  ylab("H2 ppmv") +
  xlab("Time (min)")
  
```


--- scaled plots

scale rate estimates for differences in soil quantity:

# scale estimates by soil bulk weight, and plot
```{r}
dat.combined_ed = dat.combined
dat.combined_ed$sample_id = as.character(dat.combined_ed$sample_id)
dat.combined_ed = bind_rows(dat.combined_ed, df_airdried)

dat.combined_scaled = dat.combined_ed %>%
  # set estimate =0 for PB_sand 1.5% samples since actual uptake is ~0
  mutate(estimate = ifelse(sample_id == 1.5, 0, estimate)) %>%
  # scaling for soil weight
  mutate(estimate_gSoil = estimate/soil_g)

# plots: scaled estimates for all three soils:

# k scaled to soil bulk weight vs. measured volumetric %S
dat.combined_scaled %>%
  group_by(soil_type, sample_id) %>%
  mutate(mean_k_gSoil = mean(estimate_gSoil),
         std_k = sd(estimate_gSoil),
         mean_volS = mean(volumetric_moisture_perDrySoil_rhoW),
         std_volS = sd(volumetric_moisture_perDrySoil_rhoW)) %>%
  ggplot(aes(x=mean_volS, y=mean_k_gSoil)) +
  geom_errorbar(aes(ymin = mean_k_gSoil-std_k, ymax = mean_k_gSoil + std_k, width = 0.002), color="black") +
  geom_errorbar(aes(xmin = mean_volS-std_volS, xmax = mean_volS+std_volS, width = 0.002), color="black") +
  geom_point(size=3, aes(shape = soil_type, color = soil_type)) +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  theme_linedraw() +
  ggtitle("k scaled by g-soil versus mean volumetric %S") +
  ylab("k (min-1 gSoil-1") +
  xlab("Mean volumetric %S (cm3 water-filled pores/cm3 soil)")

### FIGURE 1A
# mean %saturation vs. mean k
pdf(file =  paste0(data.path, "/data/Figure_1a_corr.pdf"), width = 6, height = 4)
dat.combined_scaled %>% 
  group_by(soil_type, sample_id) %>%
  # filter(sample_id != 7) %>%
  mutate(mean_k_gSoil = mean(estimate_gSoil),
         std_k = sd(estimate_gSoil),
         std_percSat = sd(percSat)) %>%
  ggplot(aes(x=mean_percSat, y=mean_k_gSoil)) +
  geom_errorbar(aes(ymin = mean_k_gSoil-std_k, ymax = mean_k_gSoil + std_k, width = 0.002), color="black") +
  geom_errorbar(aes(xmin = mean_percSat-std_percSat, xmax = mean_percSat+std_percSat, width = 0.002), color="black") +
  geom_point(size=2, aes(color = soil_type)) +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  theme_linedraw() +
  #ggtitle("k scaled by g-soil versus mean volumetric %S") +
  ylab("k (min-1 gSoil-1") +
  xlab("%Saturation")
dev.off()

# %saturation vs. k (replicates separately)
dat.combined_scaled %>% 
  group_by(soil_type, sample_id) %>%
  filter(sample_id != 7) %>%
  ggplot(aes(x=percSat, y=estimate_gSoil)) +
  geom_point(size=3, aes(shape = soil_type, color = soil_type)) +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  theme_linedraw() +
  ggtitle("k scaled by g-soil versus volumetric %S") +
  ylab("k (min-1 gSoil-1") +
  xlab("%Saturation")

# mean %s for molbio anal:
# dat.combined %>% select(soil_type, sample_label, mean_percSat) %>% unique()
```


# plot water potential vs. SCALED k estimates
```{r}
# log10(WP) vs. mean_k
dat.combined_scaled %>%
  group_by(soil_type, sample_id) %>%
  mutate(mean_WP = mean(log10(abs(WP))),
         std_WP = sd(log10(abs(WP))),
         mean_k_gSoil = mean(estimate_gSoil),
         std_k_gSoil = sd(estimate_gSoil)
         ) %>%
  ggplot(aes(x=mean_WP, y=mean_k_gSoil)) +
  geom_errorbar(aes(xmin = mean_WP-std_WP, xmax = mean_WP + std_WP), color="darkblue") +
  geom_errorbar(aes(ymin = mean_k_gSoil-std_k_gSoil, ymax = mean_k_gSoil + std_k_gSoil), color="darkblue") +
  geom_point(aes(color=soil_type), size=2.5) +
  theme_light() +
  xlab("Log10(water potential)") +
  ylab("Mean k (min-1 gSoil-1") +
  ggtitle("k scaled by g-soil versus Log10(WP)") +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  scale_x_reverse() +
  theme(axis.title = element_text(color="black", size=12),
        axis.text = element_text(color="black", size=12),
        legend.title = element_blank())  

# raw WP vs. mean k
dat.combined_scaled %>%
  group_by(soil_type, sample_id) %>%
  mutate(mean_WP = mean(WP),
         std_WP = sd(WP),
         mean_k_gSoil = mean(estimate_gSoil),
         std_k_gSoil = sd(estimate_gSoil)
         ) %>%
  ggplot(aes(x=mean_WP, y=mean_k_gSoil)) +
  geom_errorbar(aes(xmin = mean_WP-std_WP, xmax = mean_WP + std_WP), color="darkblue") +
  geom_errorbar(aes(ymin = mean_k_gSoil-std_k_gSoil, ymax = mean_k_gSoil + std_k_gSoil), color="darkblue") +
  geom_point(aes(color=soil_type), size=3) +
  theme_light() +
  xlab("water potential") +
  ylab("Mean k (min-1 gSoil-1") +
  ggtitle("k scaled by g-soil versus WP") +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  theme(axis.title = element_text(color="black", size=12),
        axis.text = element_text(color="black", size=11),
        legend.title = element_blank())
```

## Table S1: air-dried vs. activation thresholds
```{r}
out.path = dirname(getwd()) 
out.path
dat.combined_scaled %>%
  group_by(soil_type, sample_id) %>%
  mutate(mean_moist = mean(percSat)) %>%
  filter(sample_id %in% c(2, 3.5, 7, "air-dried")) %>%
  select(sample_id, replicate, WP, measured_percS, percSat) %>%
  distinct() 
  #write.table(file = paste0(out.path, "/data/Table1_activationThresholds.txt"), quote = FALSE, sep = "\t", row.names = FALSE)

```

### scale by OM content, SOC

Add LOI and SOC (g/g) to the df
```{r}
# add mean %LOI and sd to the df:
dat.combined.soc = dat.combined_scaled %>%
  mutate(LOI = 
           case_when(soil_type == "Forest" ~ 37.18,
                     soil_type == "Meadow" ~ 12.69,
                     soil_type == "PB_sand" ~ 3.29 ),
         LOI_sd = 
           case_when(soil_type == "Forest" ~ 1.55,
                     soil_type == "Meadow" ~ 0.22,
                     soil_type == "PB_sand" ~ 0.11 )
         )

# convert LOI to SOC (g/g)
dat.combined.soc = dat.combined.soc %>%
  mutate(SOC_gPg = (LOI/100)*0.58,
         SOC_gPg_sd = LOI_sd * (0.58/100)
         )
```

plot k (min-1 gSoil-1) vs. SOC(g/g)
```{r}
dat.combined.soc = dat.combined.soc %>%
  mutate(k_gSOC = estimate_gSoil/(SOC_gPg),
         k_gSOC_sd = k_gSOC * (SOC_gPg_sd/SOC_gPg))

# plot k (min-1 gSOC-1) vs. mean_percSat
dat.combined.soc %>% 
  group_by(soil_type, sample_id) %>%
  mutate(
    mean_k_gSOC = mean(k_gSOC),
    n_samples = n(),
    se_k_gSOC = sd(k_gSOC) / sqrt(n()),
    se_percSat = sd(percSat)/ sqrt(n())) %>%
  ggplot(aes(x=mean_percSat, y=mean_k_gSOC)) +
  geom_errorbar(aes(ymin = mean_k_gSOC-se_k_gSOC, ymax = mean_k_gSOC+se_k_gSOC), width=0.3,  color="black") +
  geom_errorbar(aes(xmin = mean_percSat-se_percSat, xmax = mean_percSat+se_percSat), width = 0.03, color="black") +
  geom_point(size=3, aes(shape = soil_type, color = soil_type)) +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  theme_linedraw() +
  ylab(expression(k ~ (min^-1 ~ SOC^-1))) +
  xlab("%Saturation")
```


Compare to max(k (min-1 gsoil-1)) vs. SOC
```{r}
dat.combined.soc %>%
  group_by(soil_type) %>%
  mutate(max_k_gSoil = max(estimate_gSoil)) %>%
  select(SOC_gPg, SOC_gPg_sd, max_k_gSoil) %>%
  # unique() %>%
  ggplot(aes(x=SOC_gPg, y=max_k_gSoil)) +
  geom_errorbar(aes(xmin = SOC_gPg-SOC_gPg_sd, xmax = SOC_gPg+SOC_gPg_sd), width = 0.003, color="black") +
  geom_point(size=3, aes(shape = soil_type, color = soil_type)) +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  theme_linedraw() +
  ggtitle("k_max (min-1 gSoil-1) versus SOC (g/g)") +
  ylab("k_max (min-1 gSoil-1") +
  xlab("SOC (g/g)")
```


### add HA-HOB relative abundance estimates to the df
### plot HA-HOB abundance vs. SOC
```{r}
dat.combined.soc = dat.combined.soc %>%
  mutate(HOB = case_when(soil_type=="Forest" ~ 34,
                         soil_type=="Meadow" ~ 24,
                         soil_type=="PB_sand" ~ 40),
         # Overall SE estimated as sqrt(sum(se^2)/n)
         HOB_err=case_when(soil_type=="Forest" ~ 0.85,
                         soil_type=="Meadow" ~ 0.79,
                         soil_type=="PB_sand" ~ 1.03))

dat.combined.soc %>%
  ggplot(aes(x=SOC_gPg, y=HOB)) +
  geom_errorbar(aes(ymin = HOB-HOB_err, ymax=HOB+HOB_err, width=0.002)) +
  geom_errorbar(aes(xmin = SOC_gPg-SOC_gPg_sd, xmax=SOC_gPg+SOC_gPg_sd, width=0.002)) +
  geom_point(aes(color=soil_type)) +
  xlab("SOC (g/g)") +
  ylab("HOB (%)") +
  ggtitle("HOB (%) vs. SOC (g/g)") 
```

### %HOB normalized by SOC(g/g) vs. max(k (min-1 gSoil-1))
```{r}
# k-max vs %HOB
dat.combined.soc %>%
  # dplyr::filter(!(soil_type == "PB_sand" & soil_g ==2)) %>%
  group_by(soil_type) %>%
  mutate(k_max_gSoil = max(estimate_gSoil)) %>%
  mutate(HOB_gSOC = HOB/(100*SOC_gPg),
         HOB_gSOC_err = HOB_gSOC * sqrt( (HOB_err/HOB)^2 + (SOC_gPg_sd/SOC_gPg)^2 ) ) %>%
  ggplot(aes(x=HOB_gSOC, y=k_max_gSoil)) +
  geom_errorbar(aes(xmin = HOB_gSOC-HOB_gSOC_err, xmax=HOB_gSOC+HOB_gSOC_err, width=0.002)) +
  geom_point(aes(color=soil_type)) +
  ylab("k-max (min-1 gSoil-1)") +
  xlab("Normalized HOB (%) / SOC (g/g)") +
  ggtitle("HOB abundance normalized by SOC")
```


### HOB abundance vs k
```{r}
dat.combined.soc %>%
  group_by(soil_type) %>%
  mutate(k_max )

dat.combined.soc %>% ungroup() %>%
  select(soil_type, estimate_gSoil, HOB, HOB_err) %>% 
  # dplyr::filter(!(soil_type == "PB_sand" & soil_g ==2)) %>%
  group_by(soil_type) %>%
  dplyr::mutate(max_k_gSoil = max(estimate_gSoil)) %>% select(-estimate_gSoil) %>%
  ggplot(aes(x=HOB, y=max_k_gSoil)) +
  geom_errorbar(aes(xmin = HOB-HOB_err, xmax=HOB+HOB_err, width=0.002)) +
  geom_point(aes(color=soil_type)) 

dat.combined.soc %>%select(soil_type, estimate_gSoil, HOB, HOB_err) %>% 
  group_by(soil_type) %>%
  dplyr::mutate(max_k_gSoil = max(estimate_gSoil),
                rel_k = estimate_gSoil/max_k_gSoil) %>%
  ggplot(aes(x=HOB, y=rel_k)) +
  geom_errorbar(aes(xmin = HOB-HOB_err, xmax=HOB+HOB_err, width=0.2)) +
  #geom_errorbar(aes(xmin = SOC_gPg-SOC_gPg_sd, xmax=SOC_gPg+SOC_gPg_sd, width=0.002)) +
  geom_point(aes(color=soil_type)) 

dat.combined.hob_relk = dat.combined.soc %>% ungroup() %>% 
  #filter(soil_type == "Forest") %>% 
  filter(sample_id != "air-dried") %>%
  select(soil_type, mean_percSat, estimate_gSoil) %>% 
  mutate(max_k_gSoil = max(estimate_gSoil), 
         rel_k = estimate_gSoil/max_k_gSoil)

hob_data = tibble(
  `%moisture` = c(6.6, 10.6, 31.9, 43.9, 55.6, 2.9, 31.2, 38.9, 77.7, 6, 16.2, 24.1, 40.4),
  HOB = c(32, 36, 34, 33, 32, 42, 39, 41, 40, 23, 26.5, 27, 20)
)

merge(dat.combined.hob_relk, hob_data, by.x = "mean_percSat", by.y = "%moisture", all.x = TRUE) %>%
  ggplot(aes(x=HOB, y=rel_k)) +
  geom_point(aes(colour = as.factor(mean_percSat)), size=2) +
  facet_wrap(~soil_type, scales="free")
```




### SOC vs. k_max:
### deduce a functional form

#### 1. quadratic fit (overfitting)
```{r}
# k vs soc
dat.combined.soc %>%
  # filter out PB_sand experiments w/ 2g soil (1-set of samples) since k-max was inferred for 4g
  dplyr::filter(!(soil_type == "PB_sand" & soil_g ==2)) %>%
  mutate(max_k_gSoil = max(estimate_gSoil)) %>%
  select(SOC_gPg, SOC_gPg_sd, max_k_gSoil) %>%
  ggplot(aes(x=SOC_gPg, y=max_k_gSoil)) +
  geom_point(aes(color=soil_type), size=3)

# try fitting a quadratic model:
soc_g = dat.combined.soc %>%
  dplyr::filter(!(soil_type == "PB_sand" & soil_g ==2)) %>% 
  dplyr::pull(SOC_gPg)

k_max_g = dat.combined.soc %>%
  dplyr::filter(!(soil_type == "PB_sand" & soil_g ==2)) %>%
  dplyr::mutate(max_k_gSoil = max(estimate_gSoil)) %>%
  dplyr::pull(max_k_gSoil)
k_max_g %>% unique()

quad_mod = lm(k_max_g ~ soc_g + I(soc_g^2))
summary(quad_mod)

coef(quad_mod)[1]

# Plot the original data
plot(soc_g, k_max_g, main = "Quadratic Fit", pch = 19, col = "blue")

# Add the fitted quadratic curve
curve(coef(quad_mod)[1] + coef(quad_mod)[2] * x + coef(quad_mod)[3] * x^2, 
      from = min(soc_g), to = max(soc_g), 
      add = TRUE, col = "red", lwd = 2)

# extract coefficients
coefs = coef(quad_mod)

intercept = coefs[1]
linear_coef = coefs[2]
quadratic_coef = coefs[3]

# plot using ggplot
data.frame(soc_g = soc_g, k_max_g = k_max_g) %>%
  ggplot(aes(x = soc_g, y = k_max_g)) +
  geom_point(color = "blue", size = 3) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), color = "red", se = FALSE, lwd = 1.5) +
  annotate("text", x = max(soc_g) * 0.7, y = max(k_max_g) * 0.8, 
           label = paste("k_max = ", round(intercept, 2), round(linear_coef, 3), "SOC(g/g) + ", round(quadratic_coef, 3), "SOC(g/g)²", sep = ""),
           color = "black", size = 4, hjust = 0.8) +
  labs(title = "Quadratic Fit", x = "SOC (g/g)", y = "k_max (min-1 gSoil-1)") +
  theme_minimal()

```

#### 2. linear fit
```{r}
dat.combined.soc %>%
  dplyr::filter(!(soil_type == "PB_sand" & soil_g ==2)) %>%
  group_by(soil_type) %>%
  mutate(k_max = max(estimate_gSoil)) %>%
  ggplot(aes(x=SOC_gPg*100, y=k_max)) +
  geom_errorbar(aes(xmin = SOC_gPg*100-SOC_gPg_sd*100, xmax=SOC_gPg*100+SOC_gPg_sd*100, width=0.002)) +
  geom_point(aes(color=soil_type), size=3) +
  geom_smooth(method="lm", se=FALSE, color="black") +
  stat_poly_eq(aes(label = paste(..eq.label.., paste0("R^2== ",..r.squared..), sep = "~~~~")),
               formula = y ~ x, 
               parse = TRUE, color="black", size=4) +
  scale_color_manual(values = c("forestgreen", "gold", "violetred")) +
  xlab("SOC %") +
  #ylab("k max (min-1 gSoil-1)") +
  ylab(expression(kmax ~ (min^-1 ~ gSoil^-1))) +
  theme_linedraw()

# linear fit: max(k) vs SOC(g/g) --> y=0.00387 + 0.24x (r-sq: 0.96)
```

### relationship between k_max and %HOB * SOC(g/g)
```{r}
# HOB*(SOC) is a proxy for the proportion of SOC that's associated with HOB

dat.combined.soc %>%
  # filter out PB_sand experiments w/ 2g soil (1-set of samples) since k-max was inferred for 4g
  dplyr::filter(!(soil_type == "PB_sand" & soil_g ==2)) %>%
  group_by(soil_type) %>%
  mutate(k_max = max(estimate_gSoil)) %>%
  ggplot(aes(y=k_max, x=(HOB/100)*SOC_gPg)) +
  geom_point(aes(color=soil_type), size=3) +
  geom_smooth(method="lm", se=FALSE, color="black") +
  stat_poly_eq(aes(label = paste(..eq.label.., ..r.squared.., sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, color="black", size=4) +
  xlab("HOB * SOC(g/g)") +
  ylab("k max (min-1 gSoil-1)") +
  ggtitle("Fraction of SOC that's HOB vs. max(k)")

# range of SOC in our dataset:
dat.combined.soc %>% 
  group_by(soil_type) %>%
  select(SOC_gPg) %>%
  unique()
```

## plot all controls
#### Fig. S2
```{r}
# read in control sample data
controls_df = read.table(paste0(data.path, "/data/H2_expt_controls_forPlotting.txt"), header=TRUE, sep="\t", check.names = FALSE)

# plot airdried samples:
airdried_plot = controls_df %>%
  filter(treatment=="air-dried") %>%
  ggplot(aes(x=Time_min, y=H2_ppm)) +
  geom_point(aes(color=as.factor(replicate), shape=Type), size=2) +
  scale_shape_manual(values = c(16,17,15)) +
  scale_color_manual(name = "Replicate", values = c("#264653", "#2A9D8F", "#E76F51")) +
  ylim(c(0,0.9)) + 
  xlab("Time (min)") +
  ylab("H2 (ppmv)") +
  theme_minimal()

# plot empty controls and autoclaved soils:
more_controls_plot = controls_df %>%
  filter(treatment!="air-dried") %>%
  ggplot(aes(x=Time_min, y=H2_ppm)) +
  geom_point(aes(color=as.factor(replicate), shape=Type), size=2) +
  scale_shape_manual(values = c(14,16,17,15)) +
  scale_color_manual(name = "Replicate", values = c("#264653", "#2A9D8F", "#E76F51", "#F4A261")) +
  ylim(c(0,0.8)) +
  xlab("Time (min)") +
  ylab("H2 (ppmv)") +
  theme_minimal()

# combine the plots:
airdried_plot + more_controls_plot +
  plot_annotation(tag_levels = 'a')
```

#### write estimates to file
```{r}
# colnames(dat.combined_scaled)
#dat.combined_scaled = dat.combined_scaled %>% select(-std.error, -init)

# the two df are in different row orders. 
# variables to merge the dfs with:
#colnames(dat.combined)
#colnames(dat.combined_scaled)
#join_vars = c("soil_type", "sample_id", "replicate", "soil_g", "measured_percS", "WP", "volumetric_moisture_perDrySoil_rhoW", "percSat", "time_min", "h2_ppmv_dil_corr")

#dat_updated = merge(dat, dat.combined_scaled, by=join_vars)
#dim(dat_updated)

#write.table(dat_updated, file=paste0(data.path, "/H2_experiments_allData_withEstimates.txt"), sep = "\t", quote = FALSE, row.names = FALSE)

```

# save image
```{r}
# data.path = dirname(getwd())
# data.path
# save.image(paste0(data.path, "/data/H2_uptakeRates_final_corr.RData"))
```

